Improving \textsl{Gaia} parallax precision \\ with a data-driven model of stars

### 2017-06-06 paper/paper.tex
dashes

In this paper we use the noise-deconvolved \cmd\ as a prior to infer more precise parallax probability distribution functions (PDFs) from \gaia\ measurements.
Understanding the validity of this method can be challenging because we are using the data both for the prior and the likelihood in our inference.
To ease understanding, here we demonstrate the general methodology using a
simpler toy model with simulated data.
Instead of inferring parallax PDFs using the 2D distribution of the \cmd\ as our prior, for this simpler model we will infer posterior PDFs of the value $Y$.
We perform this simpler inference using a prior that is also a 2D distribution: a 1D Gaussian in the y-direction, with a running mean $\mu = m\,x + b$ and some thickness $t$.
Here the thickness of the toy distribution mimics the true thickness or intrinsic scatter of the \cmd\ due to age, mass, and metallicity.
The $1\sigma$ contours as well as (noise free) samples from the toy model distribution are shown in \figurename~\ref{fig:toy}, upper-left panel.
Similarly, noisy samples from this distribution are shown in the upper-right panel.
Using a representation of the true distribution (which we learn from the noisy data) as our prior, we can infer some posterior belief over $y_{\true,n}$, a true sample from the distribution, from $y_n$, a sample from the noisy distribution.

@@

To infer the true parallax for a star using the noise-deconvolved \cmd\ as
a prior, we make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the discussion section below.
\begin{description}
\item[stationarity] We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.
  In detail, this assumption is wrong, (for example, distant halo stars are different than local
  disk stars), but the stationarity assumption is fairly weak:
  The requirement is that, locally in the \cmd, there is
  support among the high signal-to-noise stars for the types
  of stars also seen at low signal-to-noise.
  If this requirement is met, the results will not be
  strongly biased. Although this is a weak assumption, it is deep and fundamental to this
  work.
\item[selection] We have no model for the selection function of \tgas,
  nor selection volumes for any kinds of stars.
  For this reason, we are building a model of \emph{the contents of \tgas}, not the
  properties of the stars in the Milky Way (nor any volume-limited component or region
  thereof).
\item[big data] We assume that we have large numbers of stars, large enough that
  empirical Bayes, or maximum-marginalized likelihood, is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case, if a portion of the \cmd\ has only one (statistically distinct)
  exemplar---for example, a white dwarf---inference for these cases will be biased.
\item[noise model] We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the parallax uncertainty estimates,
  (the \texttt{parallax{\_}error} entry in the \gaia\ catalog), are correct.
  Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative, and leads to under-deconvolution. In other words, the features of the inferred \cmd\ will typically be broader and more conservatively estimated than than if we over-estimated the noise.
\item[dust] We treat the \psone-based three-dimensional dust maps (\citealt{green15})
  as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical Bayes model, we iteratively update each star's
  dust correction at its updated, inferred distance. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for neither
  color nor absolute magnitude.
\item[mixture of Gaussians] We assume that the \cmd\ can be represented by a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128 Gaussians.
\item[no physics] We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact).
\end{description}

@@

To generate the prior and evaluate the likelihood, we need to
correct the \tmass\ photometry for dust extinction.
With the emergence of 3D dust maps, it is now possible to estimate dust
corrections to stars within the Milky Way. The challenge is getting a
proper estimate of the distance to the star before inferring
$\varpi_{\true}$, since we need a dust correction prior to doing our
inference. Dust corrections are particularly non trivial for stars
with poor parallax signal to noise; these stars have
a likelihood that is consistent with infinite distance and can
therefore have severe dust corrections. This is most obvious in the
giant stars, which tend to have the lowest signal to noise, and can
therefore have dust corrections $> 2$ magnitudes. Although the dust
model is probabilistic, as well as our distance inference, to move
forward, we break from a fully Bayesian framework in which we might
infer the distance and the dust simultaneously (more about this in the
discussion). Instead, we correct for dust using a point estimate from
the 3D dust map \citep{green15}. We determine this dust correction by
iteratively inferring, and then sampling, our posterior PDF over
$\varpi_{\true}$. We first generate our prior using the observed,
attenuated photometry. Using this raw prior, we infer more precise
parallaxes to all the stars in our sample. We use this more
precise parallax posterior to get a measurement of dust in the 3D dust
map for each star. We then iterate this process 10 times, however
the dust values seem to converge after about 2-3 iterations. The
iterative process allows for slight variations in distance and
corresponding dust to become more consistent with the \cmd.

@@

Figure~\ref{fig:posterior} shows the various probability distributions of our inference for a few randomly-chosen objects: the likelihood (black), prior (green), and posterior PDF (blue) on the true parallaxes.
For the majority of the \gaia\ stars, our prior slightly increases the precision of the posterior PDF compared with the \gaia\ variance. Our parallax posterior PDFs are more precise than the \tgas\ catalog entries by a median factor of $1.2$. Examples of these slight increases in precision are visualized in Examples 1-4 in Figure~\ref{fig:posterior} which show slightly narrower posterior PDFs compared to the likelihoods.
However, $14\%$ of our parallax posterior PDFs are more precise than the \tgas\ catalog by a factor of $>2$. This larger change in precision is visualized in Example 5, which shows a significantly narrower posterior PDF compared to the likelihood.
The nature of our prior is also visible: it is apparent that it is made of a mixture of components, and that its impact, and shape, strongly depend on the location of each star in \cmd\ space.

@@

This change in precision is further illustrated in Figure~\ref{fig:delta}, where we show the fractional changes in the variance. The left panel shows the natural log of the fractional change in variance as a function of $(J-K_s)^C$ color, with the $1\sigma$ and $2\sigma$ contours over plotted.
The points are colored by the natural log of their fractional change in variance (the y-axis of the left panel) in both panels to help guide your eye for the right panel. The regions of \cmd\ space with the greatest improvements are shown in black, and the regions with the lowest improvements are shown in yellow.
The giants show obvious large improvements, as well as some main sequence stars.
Regions of large improvement are those with the largest photometric and parallax errors, which are the faintest observed objects.
This is a fairly standard result: the deconvolution of uncertainties is stronger in noisier regions of the data, leading to narrower features in the upper part of the \cmd, for example.
Stars with a color $(J-K_s)^C \sim 0.6$ show a decrease in precision due to the width of the prior there; the more vertical structure in the red giant branch of the \cmd.

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior.png}
\caption{Five example posterior PDFs over parallax using our method.
The upper left panel shows the position of each example data point in black, with its observed uncertainties, plotted over our prior represented by a mixture of Gaussians in green.
We chose example points that evenly spanned the color space, but within each color bin, we chose a point at random.
The remaining panels show the posterior PDF in blue, the prior PDF in green, and likelihood of the data in black.
Most stars have posterior PDFs (blue) that are only slightly more precise than the data (black), which is well represented in this random sampling of points.
Examples 1-4 are not very effected by the prior (green), where good data is good data.
Example 4 is interesting because it lies in the same color range as the red clump, where there is a lot of support from the prior, but the likelihood does not have much support there so the parallax posterior PDF, transformed into an absolute magnitude, remains on the main sequence.
Example 5 shows a significantly more precise posterior PDF compared with the likelihood, and looks very similar to the prior PDF.
It also exemplifies that our prior tends to have the largest effect on stars in the red giant branch.}
\label{fig:posterior}
\end{figure}

@@

\subsection{Distances to M67}
As a test of the accuracy of these photometric parallax inferences, we inferred the parallaxes to stars in the open cluster M67.
M67 is at a Heliocentric distance of $800$--$900~{\rm pc}$ with an estimated age of $3$--$5~{\rm Gyr}$.
Previous work \citep{cbj15} has noted that bad things happen if you try to infer a distance as the inverse of the observed (noisy) parallax when the parallax signal-to-noise (SN) is below 5, $\varpi / \sigma_\varpi < 5$.
With \gaia\ parallaxes having a minimum uncertainty of about $0.3~{\rm mas}$, the SN reaches 5 for the most certain parallax measurements at about $700~{\rm pc}$, just before the distance to this cluster.
M67 is therefore a great test case for comparing parallax inferences with or without a prior.

@@

The outputs of this project include posterior PDFs for stellar parallaxes.
It is tempting to treat these outputs as equivalent to some kind of catalog of
measurements, as we treated the original parallax measurements from \tgas.
That is, it is tempting to treat these as simply ``better'' measurements.
Taken one star at a time, this is permitted and true, and the basis for our
claim that we have de-noised the \tgas\ Catalog.
However, there are important differences, as there are with all
probabilistic catalogs (\citealt{hogg11}; \citealt{portillo17}), between the
original \tgas\ measurements and the de-noised measurements.
For users of our output, the most important difference between our
output and the \tgas\ data input is that the \tgas\ data is the representation
of a likelihood function, whereas our ouptut is a
representation of a posterior PDF, one per star.
Posterior information used naively can lead to serious statistical
errors, because each data point has had a prior multiplied in.
Multiplicative uses of the data will effectively take the prior
PDF to a large power.
We have (in previous work)
given examples of correct uses of posterior samplings or PDFs
for subsequent inferences (\citealt{hogg08}; \citealt{dfm14}); we
encourage power-users to consult those methodological
contributions before engaging with these outputs.

>>>>

In this paper we use the noise-deconvolved \cmd\ as a prior to infer more precise parallax probability distribution functions (PDFs) from \gaia\ measurements.
Understanding the validity of this method can be challenging because we are using the data both for the prior and the likelihood in our inference.
To ease understanding, here we demonstrate the general methodology using a
simpler toy model with simulated data.
Instead of inferring parallax PDFs using the 2D distribution of the \cmd\ as our prior, for this simpler model we will infer posterior PDFs of the value $Y$.
We perform this simpler inference using a prior that is also a 2D distribution: a 1D Gaussian in the $y$ direction, with a running mean $\mu = m\,x + b$ and some thickness $t$.
Here the thickness of the toy distribution mimics the true thickness or intrinsic scatter of the \cmd\ due to age, mass, and metallicity.
The $1\sigma$ contours as well as (noise free) samples from the toy model distribution are shown in \figurename~\ref{fig:toy}, upper-left panel.
Similarly, noisy samples from this distribution are shown in the upper-right panel.
Using a representation of the true distribution (which we learn from the noisy data) as our prior, we can infer some posterior belief over $y_{\true,n}$, a true sample from the distribution, from $y_n$, a sample from the noisy distribution.

@@

To infer the true parallax for a star using the noise-deconvolved \cmd\ as
a prior, we make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the discussion section below.
\begin{description}
\item[stationarity] We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.
  In detail, this assumption is wrong, (for example, distant halo stars are different than local
  disk stars), but the stationarity assumption is fairly weak:
  The requirement is that, locally in the \cmd, there is
  support among the high signal-to-noise stars for the types
  of stars also seen at low signal-to-noise.
  If this requirement is met, the results will not be
  strongly biased. Although this is a weak assumption, it is deep and fundamental to this
  work.
\item[selection] We have no model for the selection function of \tgas,
  nor selection volumes for any kinds of stars.
  For this reason, we are building a model of \emph{the contents of \tgas}, not the
  properties of the stars in the Milky Way (nor any volume-limited component or region
  thereof).
\item[big data] We assume that we have large numbers of stars, large enough that
  empirical-Bayes, or maximum-marginalized-likelihood, is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case, if a portion of the \cmd\ has only one (statistically distinct)
  exemplar---for example, a white dwarf---inference for these cases will be biased.
\item[noise model] We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the parallax uncertainty estimates,
  (the \texttt{parallax{\_}error} entry in the \gaia\ catalog), are correct.
  Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative, and leads to under-deconvolution. In other words, the features of the inferred \cmd\ will typically be broader and more conservatively estimated than than if we over-estimated the noise.
\item[dust] We treat the \psone-based three-dimensional dust maps (\citealt{green15})
  as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical Bayes model, we iteratively update each star's
  dust correction at its updated, inferred distance. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for neither
  color nor absolute magnitude.
\item[mixture of Gaussians] We assume that the \cmd\ can be represented by a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128 Gaussians.
\item[no physics] We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact).
\end{description}

@@

To generate the prior and evaluate the likelihood, we need to
correct the \tmass\ photometry for dust extinction.
With the emergence of 3D dust maps, it is now possible to estimate dust
corrections to stars within the Milky Way. The challenge is getting a
proper estimate of the distance to the star before inferring
$\varpi_{\true}$, since we need a dust correction prior to doing our
inference. Dust corrections are particularly non trivial for stars
with poor parallax signal to noise; these stars have
a likelihood that is consistent with infinite distance and can
therefore have severe dust corrections. This is most obvious in the
giant stars, which tend to have the lowest signal to noise, and can
therefore have dust corrections $> 2$~mag. Although the dust
model is probabilistic, as well as our distance inference, to move
forward, we break from a fully Bayesian framework in which we might
infer the distance and the dust simultaneously (more about this in the
discussion). Instead, we correct for dust using a point estimate from
the 3D dust map \citep{green15}. We determine this dust correction by
iteratively inferring, and then sampling, our posterior PDF over
$\varpi_{\true}$. We first generate our prior using the observed,
attenuated photometry. Using this raw prior, we infer more precise
parallaxes to all the stars in our sample. We use this more
precise parallax posterior to get a measurement of dust in the 3D dust
map for each star. We then iterate this process ten times, however
the dust values seem to converge after a few iterations. The
iterative process allows for slight variations in distance and
corresponding dust to become more consistent with the \cmd.

@@

Figure~\ref{fig:posterior} shows the various probability distributions of our inference for a few randomly-chosen objects: the likelihood (black), prior (green), and posterior PDF (blue) on the true parallaxes.
For the majority of the \gaia\ stars, our prior slightly increases the precision of the posterior PDF compared with the \gaia\ variance. Our parallax posterior PDFs are more precise than the \tgas\ catalog entries by a median factor of $1.2$. Examples of these slight increases in precision are visualized in Examples 1 through 4 in Figure~\ref{fig:posterior} which show slightly narrower posterior PDFs compared to the likelihoods.
However, $14\%$ of our parallax posterior PDFs are more precise than the \tgas\ catalog by a factor of $>2$. This larger change in precision is visualized in Example 5, which shows a significantly narrower posterior PDF compared to the likelihood.
The nature of our prior is also visible: it is apparent that it is made of a mixture of components, and that its impact, and shape, strongly depend on the location of each star in \cmd\ space.

@@

This change in precision is further illustrated in Figure~\ref{fig:delta}, where we show the fractional changes in the variance. The left panel shows the natural log of the fractional change in variance as a function of $(J-K_s)^C$ color, with the $1\sigma$ and $2\sigma$ contours over plotted.
The points are colored by the natural log of their fractional change in variance (the $y$ axis of the left panel) in both panels to help guide your eye for the right panel. The regions of \cmd\ space with the greatest improvements are shown in black, and the regions with the lowest improvements are shown in yellow.
The giants show obvious large improvements, as well as some main sequence stars.
Regions of large improvement are those with the largest photometric and parallax errors, which are the faintest observed objects.
This is a fairly standard result: the deconvolution of uncertainties is stronger in noisier regions of the data, leading to narrower features in the upper part of the \cmd, for example.
Stars with a color $(J-K_s)^C \sim 0.6$ show a decrease in precision due to the width of the prior there; the more vertical structure in the red giant branch of the \cmd.

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior.png}
\caption{Five example posterior PDFs over parallax using our method.
The upper left panel shows the position of each example data point in black, with its observed uncertainties, plotted over our prior represented by a mixture of Gaussians in green.
We chose example points that evenly spanned the color space, but within each color bin, we chose a point at random.
The remaining panels show the posterior PDF in blue, the prior PDF in green, and likelihood of the data in black.
Most stars have posterior PDFs (blue) that are only slightly more precise than the data (black), which is well represented in this random sampling of points.
Examples 1 through 4 are not very effected by the prior (green), where good data is good data.
Example 4 is interesting because it lies in the same color range as the red clump, where there is a lot of support from the prior, but the likelihood does not have much support there so the parallax posterior PDF, transformed into an absolute magnitude, remains on the main sequence.
Example 5 shows a significantly more precise posterior PDF compared with the likelihood, and looks very similar to the prior PDF.
It also exemplifies that our prior tends to have the largest effect on stars in the red giant branch.}
\label{fig:posterior}
\end{figure}

@@

\subsection{Distances to M67}
As a test of the accuracy of these photometric parallax inferences, we inferred the parallaxes to stars in the open cluster M67.
M67 is at a Heliocentric distance of $800$ to $900~{\rm pc}$ with an estimated age of $3$ to $5~{\rm Gyr}$.
Previous work \citep{cbj15} has noted that bad things happen if you try to infer a distance as the inverse of the observed (noisy) parallax when the parallax signal-to-noise (SN) is below 5, $\varpi / \sigma_\varpi < 5$.
With \gaia\ parallaxes having a minimum uncertainty of about $0.3~{\rm mas}$, the SN reaches 5 for the most certain parallax measurements at about $700~{\rm pc}$, just before the distance to this cluster.
M67 is therefore a great test case for comparing parallax inferences with or without a prior.

@@

The outputs of this project include posterior PDFs for stellar parallaxes.
It is tempting to treat these outputs as equivalent to some kind of catalog of
measurements, as we treated the original parallax measurements from \tgas.
That is, it is tempting to treat these as simply ``better'' measurements.
Taken one star at a time, this is permitted and true, and the basis for our
claim that we have de-noised the \tgas\ Catalog.
However, there are important differences, as there are with all
probabilistic catalogs (\citealt{hogg11}; \citealt{portillo17}), between the
original \tgas\ measurements and the de-noised measurements.
For users of our output, the most important difference between our
output and the \tgas\ data input is that the \tgas\ data is the representation
of a likelihood function, whereas our ouptut is a
representation of a posterior PDF, one per star.
Posterior information used naively can lead to serious statistical
errors, because each data point has had a prior multiplied in.
Multiplicative uses of the data will effectively take the prior
PDF to a large power.
We have (in previous work)
given examples of correct uses of posterior samplings or PDFs
for subsequent inferences (\citealt{hogg08}; \citealt{dfm14}); we
encourage power users to consult those methodological
contributions before engaging with these outputs.
